\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-16 ResponseToReviews.tex

2. In case of multi-region sequencing data, used for studying spatial intra-tumor heterogeneity, usually the existing tools estimate very high levels of heterogeneity. I suppose this is because they are being very stringent in terms of calling mutations especially the ones with low vafs. In that case the people end up using other tools like bamreadcount etc. to check status of these mutations in the bam files and re-estimate the extent of heterogeneity. It will be important to see how BATCAVE performs in such datasets.
\begin{response}
We agree that multi-region data are increasingly important.
In fact, the breast cancer data we test on come from a multi-region study. 
Because we built BATCAVE on top of the single-sample caller MuTect, we called variants independently across the samples in that study, and BATCAVE performed well.
Unfortunately, the sequencing in that study is not deep enough to rigorously evaluate how estimates of heterogeneity would change with and without BATCAVE.

More broadly, the BATCAVE algorithm could be implemented in a multi-regional caller that takes explicit advantage of the relationship between samples.
But such an implementation is beyond the scope of the current manuscript.

We have added a new paragraph in Discussion regarding multi-region samples, to address these issues.
\end{response}

3. Do the mutation signatures (example smoking signatures in lung cancer or UV signatures in melanoma) still appear in the more stringent BATCAVE set of mutations?
\begin{response}
In our real and simulated test cases, the mutational profile changes little between high- and low-confidence sets of mutations.
In fact, we have added a statistical test for profile change, in response to other review comments.

Approaches such as deconstructSigs can decompose an individual sample's mutational profile into a set of mutation signatures corresponding to different processes.
In practice, however, we have found these approaches to be relatively unstable.
Therefore, we do not believe it will be illuminating to apply them in the present manuscript.
\end{response}

### 2019-12-16 NAR_DRAFT.tex
push addition of tumor purity

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
\rngcomment{There are six classes of base substitution: C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all referred to by the pyrimidine of the mutated Watson-Crick pair).
This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
%Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
%At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
\bkmcomment{We make two simplifying assumptions here.
First, we assume no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
Second, we assume 100\% tumor purity.
MuTect provides a user-specified parameter allowing the user to provide an estimate of tumor purity, and downsamples reads to adjust allele frequencies accoridingly.
We encourage users to take advantage of the MuTect contamination correction where appropriate.
In the case of uncorrected contamination, actual allele frequencies will be higher than those observed, and the observed count of variants with allele frequencies between the min and max allele frequencies will be lower than the true count leading to an under-estimate of the mutation rate.
}
%We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.

@@

To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.
For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.

@@

\rngcomment{
Following MuTect \cite{Cibulskis2013}, \batcave assumes a uniform prior on the variant allele frequency.
In general, low-frequency variants are expected to be more common than high-frequency variants.
Nevertheless, the uniform frequency prior is appropriate when the goal is to make the best call possible at each potential variant site.
By favoring low-frequency and thus typically low-evidence variants, a nonuniform prior would increase the false positive rate.
Note however, that if the goal were not to best identify individual variants, but rather to best estimate the distribution of variant allele frequencies within the tumor, then a nonuniform frequency prior would be appropriate \cite{Nielsen2012}.
}

>>>>

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
\bkmcomment{There are six classes of base substitution -- C$>$A, C$>$G, C$>$T, T$>$A, T$>$C, T$>$G (all substitutions are referred to by the pyrimidine of the mutated Watson - Crick base pair).
This results in a total of 32 ($4\times 2 \times 4$) reference contexts and 3 alternate bases per reference base.}
%Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
%At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
\bkmcomment{Low tumor purity can result in a distribution of reported allele frequencies that differs from the true distribution.
We suggest in cases where an estimate of tumor purity is known, such as when using the GATK 4 best practices pipeline, providing it to batcaver.}
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ and tumor_purity are free parameters.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.

@@

To improve variant identification, the context-dependent prior probability of mutation must converge to an accurate representation of the data generating distribution within the set of high-confidence mutations.
When applied to simulated data, the prior converged within a few hundred mutations (Fig.~\ref{NAR-kl_fig}), \bkmcomment{and the \texttt{batcaver} package emits a warning if there are fewer than 500 high-confidence mutations.}.
For comparison, in our simulated data sets the number of high-confidence mutations ranged between 1,500 and 5,000, and in the real AML we test on it is over 17,000~\cite{Griffith2015}.

@@
